---
date: "`r format(Sys.time(), '%Y-%m-%d, %H:%M')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
params:
  cytob.id:
    label: Cytobank ID
    value: NULL
    input: text
  spade.id:
    label: SPADE ID, if SPADE clustering available
    value: NULL
    input: text
  live.gate:
    label: Gate used to gate to live cells
    value: NULL
    input: text
  fcs.dir:
    label: Path to the directory of processed data.
    value: NULL
    input: text
  spade.suffix:
    label: Regular expression of suffix to strip from FCS filenames after SPADE
    value: NULL
    input: text
  annotation: 
    label: Path to sample annotation
    value: "annotation.txt"
    input: file
  markers: 
    label: Path to marker annotation
    value: "markers.txt"
    input: file
  gating: 
    label: Path to gating XML
    value: "gating.xml"
    input: file
  latest.gating: 
    label: Path to gating XML
    value: NULL
    input: file
  mst: 
    label: Path to minimum spanning tree GML
    value: "mst.gml"
    input: file
  bubbles: 
    label: Path to bubbles RDS
    value: "bubbles.rds"
    input: file
  metas: 
    label: Path to metacluster RDS
    value: "metas.rds"
    input: file
  info: 
    label: Path to processing information YAML
    value: "info.yaml"
    input: file
  author:
    label: Author to include in the HTML report
    value: ""
    input: text
  stringsAsFactors:
    label: Convert strings to factor when reading in flat files?
    value: FALSE
    input: checkbox
  overwrite:
    label: Overwrite any CDF or RDS files with the same name and file path as this Rmd?
    value: FALSE
    input: checkbox
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_file = paste0(xfun::sans_ext(input), '.html'),
      envir = globalenv()
    )
  })    
---

---
title: "Experiment `r as.character(params$cytob.id)`: Create `cytofData` Object`r if (!is.null(params$spade.id)) ' after SPADE Clustering'`"
author: "`r params$author`"
---

```{r setup, include = FALSE}
## chunk must be called setup for chunk options to be respected in html_notebook
library(knitr)
opts_chunk$set(message = FALSE, comment = "")
library(BiocGenerics)
```

```{r shiny-setup, include = FALSE}
## setup shiny progress bar
progress <- requireNamespace("shiny", quietly = TRUE) && shiny::isRunning()
n_progress <- 9
```


```{r info-lines, results = "asis", echo = FALSE}
if (is.null(params$cytob.id)) stop("Cytobank ID must be specified")
spade <- !is.null(params$spade.id)
if (is.null(params$live.gate)) stop("Live gate must be specified")
cat("__Cytobank ID__: ", as.character(params$cytob.id), "  \n", # need two spaces at eol
    if (spade) paste("__SPADE ID__: ", as.character(params$spade.id), "  \n"),
    "__Live gate__: ", as.character(params$live.gate), sep = "")
```

# Load Packages

This notebook creates an R6 cytofData object from processed CyTOF data and 
optionally the results of cell clustering (e.g. via SPADE). We require 
the **ncdfFlow** package to work with ncdfFlowSets, **data.table** to read in 
flat files (`.csv` or `.tsv`), **yaml** to read in YAML and **igraph** to read 
in the minimum spanning tree from SPADE clustering, if available. The 
**CytoML**, **flowWorkspace**, and **XML** packages are required to obtain 
gating information. Finally **tibble** is used to display data frames, while
**sessioninfo** and **tools** are used to handle session information and file 
paths respectively.

```{r packages, message = FALSE}
library(cytofR6)
required <- c("rmarkdown", "data.table", "yaml", "igraph"[spade],
              "CytoML", "flowWorkspace", "XML",
              "tibble", "sessioninfo", "tools")
required_available <- sapply(required, require, quietly = TRUE, 
                             character.only = TRUE)
if (!all(required_available)) 
  stop("The following packages must be installed to use the Rmd template\n", 
       paste(required[!required_available], collapse = ", "))
rm(required, required_available)
```

```{r session-info, include = FALSE}
sess_info <- sessioninfo::session_info()
```


# Setup 

```{r progress-setup, include = FALSE}
if (progress)
  shiny::incProgress(1/n_progress, detail = "Setup")
```

We check that the directory containing the FCS files has been specified and 
define a directory for the outputs.

```{r dir-setup}
if (is.null(params$fcs.dir)) stop("FCS directory must be specified")
# use fig.path to get user-specified output file & dir from `render()` call
output.dir <- dirname(dirname(opts_chunk$get()$fig.path))
# if dir not specified, use directory where Rmd is (knit working directory)
if (output.dir == "."){
  output.dir <- knitr::opts_knit$get("output.dir")
  # if running interactively, use getwd() (= knit wd from RStudio)
  if (is.null(output.dir)) output.dir <- getwd()
}
output.file <- sub("([^_]*)_files", "\\1", 
                   basename(dirname(opts_chunk$get()$fig.path)))
# check whether saving `cytofData` will overwrite existing files
nc_exists <- file.exists(file.path(output.dir, paste0(output.file, ".nc")))
rds_exists <- file.exists(file.path(output.dir, paste0(output.file, ".rds")))
if (!params$overwrite && (nc_exists | rds_exists)){
  stop("`params$overwrite` is `FALSE` yet ", output.dir, 
       " contains:\n",
       paste0(output.file, ".nc\n")[nc_exists],
       paste0(output.file, ".rds\n")[rds_exists],
       "Set `params$overwrite` to `TRUE` or change name/file path of .Rmd")
}
kable(data.frame(Name = c("FCS directory", "Output directory"),
                 Path = c(params$fcs.dir, output.dir)))
rm(rds_exists)
```

# Validating inputs

```{r progress-validate, include = FALSE}
if (progress)
  shiny::incProgress(1/n_progress, detail = "Validating inputs")
```

Now we validate the latest gating file (if provided) and the sample annotation.

## Latest gating file

If an updated gating file has been specified, we check the live gate in the 
this file against that used when processing the data - if the parameters 
(defining channels) or boundaries have changed, the notebook will terminate.

```{r check-gating}
if (!is.null(params$latest.gating)){
  if (md5sum(params$gating) != md5sum(params$latest.gating)){
    # latest gating
    doc <- xmlParse(params$latest.gating)
    # get names of Boolean gates
    gates <- xpathSApply(doc, "//gating:BooleanGate/descendant::name",
                         xmlValue)
    # id the one corresponding to the live gate
    id <- which(gates == as.character(params$live.gate))
    if (!length(id)) stop("Cannot find live gate in latest gating file.")
    # get the IDs for gates contributing to live gate
    gate_id <- xpathSApply(doc, 
                           paste0("//gating:BooleanGate[", id,
                                  "]/gating:and/gating:gateReference"),
                           xmlGetAttr, "gating:ref")
    # get the XML nodes corresponding to these gate IDs
    nodes <- getNodeSet(doc, paste0("//*[@gating:id='", gate_id, "']", 
                                    collapse = " | "))
    names(nodes) <- xpathSApply(doc, paste0("//*[@gating:id='", gate_id,
                                            "']/descendant::name", 
                                            collapse = " | "), xmlValue)
    
    # old gating
    doc0 <- xmlParse(params$gating)
    # get names of Boolean gates
    gates0 <- xpathSApply(doc0, "//gating:BooleanGate/descendant::name",
                          xmlValue)
    # id the one corresponding to the live gate
    id0 <- which(gates0 == as.character(params$live.gate))
    if (!length(id0)) stop("Cannot find live gate in gating file.")
    # get the IDs for gates contributing to live gate
    gate_id0 <- xpathSApply(doc0, 
                            paste0("//gating:BooleanGate[", id0,
                                   "]/gating:and/gating:gateReference"),
                            xmlGetAttr, "gating:ref")
    # get the XML nodes corresponding to these gate IDs
    nodes0 <- getNodeSet(doc0, paste0("//*[@gating:id='", gate_id0, "']", 
                                      collapse = " | "))
    names(nodes0) <- xpathSApply(doc0, paste0("//*[@gating:id='", gate_id0,
                                              "']/descendant::name", 
                                              collapse = " | "), xmlValue)
    
    # compare gates
    field.ok <- FALSE
    if (setequal(names(nodes), names(nodes0))){
      for (nm in names(nodes)){
        field <- setdiff(names(nodes[[nm]]), c("custom_info"))
        field.ok <- all.equal(nodes[[nm]][field], nodes0[[nm]][field])
        if (!field.ok) break
      }
    }
    if (!field.ok) 
      stop("The live gate has changed since the CyTOF data were processed. ",
           "Please extract features with the latest gating file.")
    rm(gates, id, gate_id, nodes, gates0, id0, gate_id0, nodes0, 
       field, field.ok, nm)
    free(doc)
    free(doc0)
    rm(doc, doc0)
  }
  gate.file <- params$latest.gating
} else gate.file <- params$gating
```

## Sample annotation

We read in the sample annotation:

```{r sample-annotation}
ext <- file_ext(params$annotation)
if (tolower(ext) == "rds"){
  samples <- readRDS(params$annotation)
} else {
  samples <- fread(params$annotation, 
                   data.table = FALSE,
                   header = TRUE,
                   stringsAsFactors = params$stringsAsFactors)
}
rm(ext)
as_tibble(samples)
```

## Input files

We match the names of the FCS file to the sample annotation - 
if SPADE clustering has been applied, the suffix added to the FCS filenames 
after clustering is removed before matching. Only annotated FCS files will be 
included in the `cytofData` object.

```{r match-annotation}
input.files <- dir(params$fcs.dir, pattern = ".fcs$")
if (spade){
  if (!file.exists(params$mst)) 
    stop(params$mst, " file does not exist")
  spade.suffix <- params$spade.suffix
  if (is.null(spade.suffix)) {
    # escape special characters in live gate
    tmp.live.gate <- gsub("([[:punct:]])", "\\\\\\1", 
                          as.character(params$live.gate))
    spade.suffix <- paste0("_", tmp.live.gate, ".*")
  }
  input.files <- grep(spade.suffix, input.files, value = TRUE)
  id <- match(samples[[1]],
              sub(paste0("(.*)", spade.suffix), "\\1.fcs", input.files), 
              nomatch = 0L)
} else {
  id <- match(samples[[1]], input.files, nomatch = 0L)
}
# Throw error if annotated but not available
if (any(id == 0L))
  stop("Could not match first column of sample annotation to input files in\n",
       params$fcs.dir,
       if (spade)
         paste0("\n\nRegular expression for suffix to strip from ",
                "FCS file names:\n", spade.suffix),
       "\n\nAnnotated samples with no match:\n",
      paste(samples[[1]][id == 0L], collapse = "\n"))
# Keep annotated FCS files only, drop note to user if filtering
if (length(input.files) > length(!is.na(id)))
  message("The following samples are excluded by the sample annotation file:\n",
          paste(input.files[-id], collapse = "\n"))
if (spade) rm(spade.suffix)
tibble(`Input file` = input.files[id])
```

# Data structures

Now we prepare the data structures required to create the `cytofData` object.

## ncdfFlowSet

```{r progress-readFCS, include = FALSE}
if (progress)
  shiny::incProgress(1/n_progress, detail = "Read in FCS files")
```

An `AnnotatedDataFrame` is created from the sample annotation, which is used 
as the `phenoData` for a `ncdfFlowSet`:

```{r read-FCS}
annots <- as(cbind(data.frame(name = samples[[1]], 
                              row.names = samples[[1]],
                              stringsAsFactors = params$stringsAsFactors),
                   samples[, -1]),
             "AnnotatedDataFrame")
if (params$overwrite || !nc_exists){ # only needed for interactive use
  inputSet <- suppressMessages(
    read.ncdfFlowSet(file.path(params$fcs.dir, input.files[id]), 
                     phenoData = annots, 
                     ncdfFile = file.path(output.dir, output.file))
  )
} else stop ("`params$overwrite` is `FALSE` and CDF file already exists")
rm(annots, id, input.files, nc_exists, samples)
inputSet
```

## Marker annotation

The processed marker annotation is read in from the specified file:

```{r marker-annot}
markers <- fread(params$markers, 
                 data.table = FALSE,
                 header = TRUE,
                 stringsAsFactors = params$stringsAsFactors)
if (!all(c("name", "desc") %in% names(markers)))
  stop("Please review the marker annotation file. It should include ",
       "the columns 'name' and 'desc'.")
as_tibble(markers)
```

## Populations

```{r progress-gating, include = FALSE}
if (progress)
  shiny::incProgress(1/n_progress, detail = "Get gating information")
```

Gating information is obtained from the latest gating file available.

```{r gating}
suppressMessages(treeML <- read.gatingML.cytobank(gate.file))
gatingSet <- GatingSet(inputSet)
suppressMessages(gating(treeML, gatingSet))
rm(inputSet, treeML)
```

We extract the corresponding population assignments.

```{r progress-populations, include = FALSE}
if (progress)
  shiny::incProgress(1/n_progress, 
                     detail = "Extract latest population assignments")
```

```{r populations}
nodes <- getNodes(gatingSet)
populations <- lapply(gatingSet, function(gh){
  cur.pops <- lapply(nodes, function(nd) getIndiceMat(gh, nd))
  do.call(cbind, cur.pops)
})
populations <- do.call(rbind, populations)
rm(nodes)
str(populations)
```

## Bubbles

```{r progress-bubbles, include = FALSE}
if (progress)
  shiny::incProgress(1/n_progress, detail = "Load and match bubbles")
```

If provided, bubbles are loaded and matched to cluster annotations.

```{r bubbles, warning = FALSE}
if (file.exists(params$bubbles)){
  bubbles <- readRDS(params$bubbles)
  clusters <- fsApply(getData(gatingSet)[, "cluster"], exprs)
  bubbles <- lapply(bubbles, function(bub) clusters %in% unlist(bub))
  bubbles <- do.call(cbind, bubbles)
  if(!file.exists(params$metas)) {
    rm(clusters)
  }
  str(bubbles)
} else bubbles <- NULL
```

## Metaclusters

```{r progress-metas, include = FALSE}
if (progress)
  shiny::incProgress(1/n_progress, detail = "Load and match bubbles")
```

If provided, metaclusters are loaded and matched to cluster annotations.

```{r metas, warning = FALSE}
if (file.exists(params$metas)){
  metas <- readRDS(params$metas)
  if(!"clusters" %in% ls()) {
    clusters <- fsApply(getData(gatingSet)[, "cluster"], exprs)
  }
  metas <- lapply(metas, function(bub) clusters %in% unlist(bub))
  metas <- do.call(cbind, metas)
  rm(clusters)
  str(metas)
} else metas <- NULL
```

## Obtain Minimum Spanning Tree

```{r progress-mst, include = FALSE}
if (progress)
  shiny::incProgress(1/n_progress, detail = "Obtain minimum spanning tree")
```

The minimum spanning tree is read in, if available.

```{r mst}
if (spade){
  ## tree structure
  mst <- read.graph(params$mst, format ="gml")
  summary(mst)
} else {
  mst <- NULL
}
```

## Update processing information

```{r progress-info, include = FALSE}
if (progress)
  shiny::incProgress(1/n_progress, detail = "Update processing information")
```

We collate metadata on creating the `cytofData` object. If information has been 
provided on previous processing, this is information is added as a separate 
section:

```{r info}
# previous info
if (file.exists(params$info)){
  info <- read_yaml(params$info)
} else info <- NULL
# new info
new_info <- list(
  "Path to processed FCS files" = params$fcs.dir,
  "Gating file (populations)" = gate.file,
  "Sample annotation" = params$annotation,
  "Marker annotation" = params$markers,
  "Clustering" = ifelse(spade, "SPADE", "none"),
  "Notebook version" = as.character(packageVersion("cytofR6")),
  "Date" = format(Sys.Date(), "%Y-%m-%d"))
if (is.character(unlist(info, recursive = FALSE))){
  # only one level of YAML
  info <- list(Processing = info, "Creating cytofData object" = new_info)
} else {
  info <- c(info, list("Creating cytofData object" = new_info))
}
rm(gate.file, new_info, spade)
cat(as.yaml(info))
```

# Save results

```{r progress-save, include = FALSE}
if (progress)
  shiny::incProgress(1/n_progress, detail = "Create and save R6 object")
rm(n_progress, progress)
```

We create a cytofData R6 object containing the processed data, annotation, 
population assignments, cell clustering and processing information. The result 
is saved in the output directory: `r output.dir`, as the files listed below:

```{r cytofData}
## create R6 object
cytof <- cytofData$new(cytofSet = getData(gatingSet),
                       marker_data = markers,
                       populations = populations,
                       mst = mst,
                       bubbles = bubbles,
                       metaclusters = metas,
                       info = info)
cytof$save(output.file, cdf = "skip", dir = output.dir, 
           overwrite = params$overwrite)
tab <- data.frame(File = c(paste0(output.file, ".nc"), 
                          paste0(output.file, ".rds")),
                 Description = c("On disk cdf storage for processed CyTOF data",
                                 "R6 object of processed CyTOF data"))
rm(gatingSet, markers, populations, mst, bubbles, info, 
   output.file, output.dir)
kable(tab)
```

# Addendum

This is notebook version `r as.character(packageVersion("cytofR6"))`.

Platform information:
```{r session-platform, warning = FALSE, echo = FALSE}
kable(data.frame(configuration = names(sess_info$platform), 
                 value = unlist(sess_info$platform),
                 row.names = NULL))
```

R packages used: 
```{r session-packages, warning = FALSE, echo = FALSE}
keep <- c("package", "ondiskversion", "date", "source")
tab <- as_tibble(unclass(sess_info$packages)[keep])
colnames(tab)[2] <- "version"
rm(sess_info, keep)
kable(tab)
```

```{r clean-up, echo = FALSE}
rm(tab)
```
  
